Interpretable Neural Networks using EAGGA

### 2025-03-13 paper/main.tex
add bullets here and there

      * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation
        (implicit in the loss function for better numerical stability, NN itself outputs "logits", which refers to pre-activation output in pytorch)

@@

    - then evaluate on last fold from CV + keep best $\mu$ individuals + generate $\lambda=10$ offspring for new generation + repeat

@@

@@

\section{Experimental Results and Discussion}
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{../code/export/plot_test_dhvs.png}  % scale figure, cf. https://tex.stackexchange.com/a/16584
  \caption{caption}
  \label{fig-test-dhvs}
\end{figure}

\section{Conclusion and Future Outlook}
- MO BO on group structure space possible?
- here is a citation \cite{EAGGA}.

@@

\section{}
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{../code/export/plot_early_stopping_losses_pareto_set.png}
  \caption{caption}
  \label{fig-es-losses}
\end{figure}

\section{}
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{../code/export/plot_val_dhvs_over_time.png}
  \caption{caption}
  \label{fig-val-dhvs}
\end{figure}

>>>>

      * hidden layers use ReLU activation and dropout afterwards, then a shared output layer with concatenated sub-NNs' outputs as input and sigmoid activation with
        (sigmoid implicit in the loss function for better numerical stability, NN itself outputs "logits", which refers to pre-activation output in pytorch)
      * loss is binary cross entropy loss, optimizer is AdamW with default params

@@

    - then evaluate on last remaining from CV + keep best $\mu$ individuals + generate $\lambda=10$ offspring for new generation + repeat
    - final evaluation of individuals of pareto set
      * for each individual, train model on training set of hold-out for max \# of epochs taken during CV-training
      * decided on max \# epochs instead of mean after looking at loss graphs in preliminary experiments, also refer Figure \ref{fig-es-losses}
      * those show that there is no uptick in losses on the early stopping portion (disjunct from training portion) of the set, hence max reasonable

@@

  * did not train on philippine and gina datasets (308, 970 features, respectively), as they ran out of memory when computing interaction detectors

@@

\section{Experimental Results and Discussion}
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{../code/export/plot_test_dhvs.png}  % scale figure, cf. https://tex.stackexchange.com/a/16584
  \caption{caption}
  \label{fig-test-dhvs}
\end{figure}
- preliminary
  * our implementation of feature + interaction detectors likely tend to include
    - more features + interactions for high p datasets
    - less for low p datasets
    - remember: original paper samples \# of features included from truncated geometric distribution, what is not mentioned is probability of this distribution
      is determined from fitting 10 trees + looking at relative \# of features used (cf. their github repo ~/R/TunerEAGGA.R function get\_n\_selected\_rpart())
    -> mlr3 default decision tree max depth 30
      * i.e. datasets with p < 30 likely use all features, which translates to trunc geom prob = 1
      * vice-versa for p >> 30 datasets relative \# features < 0.5 (our trunc geom prob)
      * similar reasoning for pairwise interactions
  * individuals not in Pareto sets occassionally exhibit AUC < 0.5
    - this is not by mistake, usually AUC on binary target can simply be inverted (simply predict the opposite class)
    - this cannot easily be inverted due to monotonicity constraints (weights cannot simply be multiplied by -1 if constraint)
    - could happen e.g. due to AdamW and weight clipping (employed to enforce monotonicity)
      * AdamW uses momentum
      * possible scenario: large momentum would yield negative weights, but after epoch they're clipped to [0,infty) + training stops due to early stopping
      * weight clipping very imperfect way of enforcing monotonicity, but currently in pytorch unfortunately only way to implement this
- Figure \ref{fig-test-dhvs} suggests comparable performance to XGBoost EAGGA
- overview over pareto sets can be accessed on our github repo at ~/code/export/*.csv
  * NN HPs
    - total layers mostly 3-4, most commonly 3, goes as high as 7 (Satellite (36), diabetes (8))
    - nodes per hidden layer mostly 3-6, goes as high as 12 (diabetes (8))
    - p dropout goes as high as 0.7 (climate (18), spambase (57)), but mostly in the 0.1 to 0.3 range
  * group structures
    - great diversity in NF across datasets
      * phoneme (5) up to 1
      * blood (4) up to 1
      * banknote (4) up to 0.75
      * diabetes (8) up to 0.5
      * climate (18) up to 0.67
      -> low p datasets in the benchmark tend to have higher NF
        - possible consequence of shorter evaluation time
        - low p datasets have much higher \# generations
        - group structure space much more likely to be exhausted, i.e. more exploration of NN hps
    - NI, NNM consequently (bounded by NF, can never be more than max \# for respective \# included features) rather low low
  * dhv contributions
    - predominately low for fitted models
    - mostly highest for featureless learner predicting majority class
    -> sign of good exploration of pareto front + stable estimate, refer Section \ref{sec-moo-post}
- loss graphs Figure \ref{fig-es-losses}
  * suggests models could have benefitted from longer training on some datasets, as some loss curves haven't converged when stopping criterion hit
  * crit was likely triggered by short-term spike, thus could potentially be resolved by comparing average of last k loss against average of >patience< losses prior
    to that to not be as exposed to short-term spikes in loss
  * on other datasets, graphs suggest earlier stop would have been totally fine as the networks have long converged, but longer training likely not an issue as loss
    graphs come from early stopping dataset portion, which is disjunct from training set
- dhv over generations Figure \ref{fig-val-dhvs}
  * compute on val sett, i.e. had 5 folds per individual
    - NF, NI, NNM always the same for each fold
    - but 5 different AUCs
    -> hypervolume of (mean(AUC 1, AUC 2, ..., AUC 5), NF, NI, NNM)
  * artifacts / drops along y likely due to inconsistent computation of Pareto front by third-party library
    - preliminary experiments on dummy pareto fronts with 10x4 metrics: noticed nds function returning different rankings for same front
    - returned ranking switched back and forth between two only slightly different options (only 1 or 2 indices were swapped)
    - not sure what caused this, as made same observation using another library that was planned as alternative
    - thus unfortunately not fixable for me
  * all but 3 datasets (2 of which only trained for 1 generation, aynway) show improvement of dominated hypervolume over generations
  * BUT: absolute as well as relative improvement almost negligible -> no large final dhv decrease would we just have evaluated the models gotten from the detectors
    - this was also the reason we didn't run EAGGA on philippine and gina simply without the detectors (as that's where they crashed) -> evidence suggested that detectors are vital to initial performance
    - original paper supports this assumption \citep[Fig. 4, p. 545]{EAGGA}

\section{Conclusion and Future Outlook}
- MO BO on group structure space possible?

@@

\section{Plots}
\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{../code/export/plot_early_stopping_losses_pareto_set.png}
  \caption{Pareto set loss of all datasets, evaluated on early stopping set. Same colours denote losses coming from folds of the same individual.
            x-axis portrait epochs, y-axis binary cross entropy loss.}
  \label{fig-es-losses}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.9\linewidth]{../code/export/plot_val_dhvs_over_time.png}
  \caption{Dominated hypervolume over generations, evaluated on validation set.
            x-axis portrait generations, y-axis dominated hypervolume using mean AUC over folds.}
  \label{fig-val-dhvs}
\end{figure}
